---
title: "Drinking Vessels, Skeletal Remains, and Hair Pins Analysis"
subtitle: "Laconia and Messinia Burial Data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: inline
---

## Introduction

This document presents the quantitative analyses underlying the thesis. The dataset comprises funerary contexts from Laconia and Messenia spanning the Middle Helladic through Late Helladic periods (approximately 2000–1070 BCE). Each record corresponds to a single tomb or grave feature and includes information on tomb type, chronological attribution, burial practice, and the presence of drinking vessels, skeletal remains, and hair pins.

The analyses address four main questions: (1) how tomb types and burial practices are distributed across the two regions; (2) whether drinking vessels and skeletal remains co-occur in patterned ways; (3) whether hair pins cluster with particular burial practices, vessel assemblages, or tomb sizes; and (4) how these variables change over the course of the Bronze Age. All code is folded by default and can be expanded for inspection.

## Load Libraries

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(janitor)
library(knitr)
library(kableExtra)
library(broom)
library(ggrepel)
```

## Visualization Settings

```{r color-palette}
# Set your color palette here
colors <- list(
  laconia = "#1B9E77",      # Laconia color (default: teal/green)
  messenia = "#D95F02",     # Messenia color (default: orange)
  drinking_vessel = "#7570B3",  # Drinking vessels (default: purple)
  skulls = "#E7298A",        # Skulls (default: pink)
  pin = "#E6AB02",            # Hair pins (default: gold/amber)
  background = "#FFFFCC"     # Background/neutral (default: light yellow)
)

# These colors will be used throughout all plots
# Change the hex codes to your preferred colors

# Set global ggplot theme with larger text
theme_set(theme_minimal(base_size = 20))
theme_update(
  plot.title = element_text(size = 22, face = "bold"),
  plot.subtitle = element_text(size = 18),
  axis.title = element_text(size = 18),
  axis.text = element_text(size = 16),
  legend.text = element_text(size = 16),
  legend.title = element_text(size = 17),
  strip.text = element_text(size = 17)
)
```

## Data Source

```{r load-data}
# Load the CSV data
data_path <- here::here("data", "laconia_messinia_cups_skulls_pins.csv")
cups_skulls <- read_csv(data_path)

# Display dimensions
cat("Dataset dimensions:", nrow(cups_skulls), "graves ×", ncol(cups_skulls), "columns\n")

# Show column names
glimpse(cups_skulls)
```

## Data Preparation

### Variable Coding

Raw text fields were recoded into analysis-ready variables. Drinking vessel presence, skeletal remains, and hair pin presence were each recorded as binary yes/no fields in the spreadsheet and converted to logical values (`TRUE`/`FALSE`); entries that could not be classified were retained as `NA` and excluded from relevant calculations. Primary and secondary burial were coded from "present"/"absent" fields in the same way.

Burial contexts were then classified into four mutually exclusive categories based on the combination of primary and secondary burial evidence: *Primary Only*, *Secondary Only*, *Both Primary and Secondary*, and *Unknown/Not Recorded*. Tomb types were simplified from their original descriptive labels into five structural categories — Tholos, Chamber Tomb, Cist Grave, Pit Grave, and Dromos Only — to allow comparison across sites where nomenclature varies.

```{r clean-data}
cups_skulls_clean <- cups_skulls %>%
  # Clean column names
  janitor::clean_names() %>%

  # Trim whitespace from character columns
  mutate(across(where(is.character), str_trim)) %>%

  # Standardize yes/no values in drinking vessel column
  mutate(
    has_drinking_vessel = case_when(
      tolower(drinking_vessel) == "yes" ~ TRUE,
      tolower(drinking_vessel) == "no" ~ FALSE,
      TRUE ~ NA
    ),
    has_remains = case_when(
      tolower(remains) == "yes" ~ TRUE,
      tolower(remains) == "no" ~ FALSE,
      TRUE ~ NA
    ),
    primary_burial = case_when(
      tolower(primary) == "present" ~ TRUE,
      tolower(primary) == "absent" ~ FALSE,
      TRUE ~ NA
    ),
    secondary_burial = case_when(
      tolower(secondary) == "present" ~ TRUE,
      tolower(secondary) == "absent" ~ FALSE,
      TRUE ~ NA
    ),
    has_pin = case_when(
      tolower(hair_pin) == "yes" ~ TRUE,
      tolower(hair_pin) == "no" ~ FALSE,
      TRUE ~ NA
    )
  ) %>%

  # Create burial category
  mutate(
    burial_category = case_when(
      primary_burial & secondary_burial ~ "Both Primary & Secondary",
      primary_burial & !secondary_burial ~ "Primary Only",
      !primary_burial & secondary_burial ~ "Secondary Only",
      TRUE ~ "Unknown/Not Recorded"
    )
  ) %>%

  # Simplify grave type
  # ** YOU CAN CUSTOMIZE THIS SECTION **
  # If you want different groupings, edit the patterns here
  # Current grouping is based on archaeological complexity
  mutate(
    grave_type_simple = case_when(
      str_detect(tolower(type), "tholos") ~ "Tholos",
      str_detect(tolower(type), "chamber") ~ "Chamber Tomb",
      str_detect(tolower(type), "cist") ~ "Cist Grave",
      str_detect(tolower(type), "pit") ~ "Pit Grave",
      str_detect(tolower(type), "dromos") ~ "Dromos Only",
      TRUE ~ "Other"
    )
  )

# Show cleaned data structure
glimpse(cups_skulls_clean)
```

### Chronological Classification

Period strings as recorded in the original dataset (e.g., "LHIIA," "LHI–LHIII") were matched against a period dictionary that assigns each code an earliest and latest component phase. Calendar date ranges were then attached by joining against a lookup table of BCE start and end dates for each component phase, based on conventional Aegean Bronze Age chronology. The period dictionary and calendar date table are loaded and applied below.

```{r load-dictionaries, eval=TRUE}
# NOTE: Set eval=TRUE once the student has completed the dictionaries

# Load period dictionary
period_dict <- read_csv(here::here("data", "period_dictionary.csv")) %>%
  mutate(across(where(is.character), str_trim))

# Load tomb type dictionary
tomb_dict <- read_csv(here::here("data", "tomb_type_dictionary.csv")) %>%
  mutate(tomb_type = str_trim(tomb_type))

# Join period dictionary and extract period components
cups_skulls_clean <- cups_skulls_clean %>%
  left_join(
    period_dict %>% select(period_code, chronological_order, earliest_component, latest_component),
    by = c("period" = "period_code")
  ) %>%

  # Get ordinal values for earliest component
  left_join(
    period_dict %>% select(period_code, chronological_order) %>% rename(start_order = chronological_order),
    by = c("earliest_component" = "period_code")
  ) %>%

  # Get ordinal values for latest component
  left_join(
    period_dict %>% select(period_code, chronological_order) %>% rename(end_order = chronological_order),
    by = c("latest_component" = "period_code")
  ) %>%

  # For single periods (where start = end), use that value
  # For ranges, create midpoint
  mutate(
    period_start_order = start_order,
    period_end_order = end_order,
    period_midpoint_order = (start_order + end_order) / 2
  )

# Calendar year lookup for each single-component period code
# Sources: user-supplied dates; sub-phases (MHIIIB, LHIIIB1, LHIIIC1) estimated
component_cal <- tribble(
  ~component,      ~cal_start_bce, ~cal_end_bce,
  "MH",            2000,           1600,
  "MHII",          1850,           1700,
  "MHIII",         1700,           1600,
  "MHIIIB",        1650,           1600,
  "LH",            1600,           1070,
  "LHI",           1600,           1500,
  "LHII",          1500,           1390,
  "LHIIA",         1500,           1440,
  "LHIIB",         1440,           1390,
  "LHIII",         1390,           1070,
  "LHIIIA",        1390,           1300,
  "LHIIIA1",       1390,           1370,
  "LHIIIA2",       1370,           1300,
  "LHIIIB",        1300,           1190,
  "LHIIIB1",       1300,           1250,
  "LHIIIC",        1190,           1070,
  "LHIIIC1",       1190,           1130,
  "submycenaean",  1070,           1015
)

# Add calendar year columns: start = start of earliest component,
# end = end of latest component
cups_skulls_clean <- cups_skulls_clean %>%
  left_join(
    component_cal %>% select(component, cal_start_bce),
    by = c("earliest_component" = "component")
  ) %>%
  rename(period_start_cal = cal_start_bce) %>%
  left_join(
    component_cal %>% select(component, cal_end_bce),
    by = c("latest_component" = "component")
  ) %>%
  rename(period_end_cal = cal_end_bce) %>%
  mutate(period_midpoint_cal = (period_start_cal + period_end_cal) / 2)

# Broad phase groupings with calendar spans for temporal plots
phase_cal <- tribble(
  ~broad_phase,  ~phase_start_bce, ~phase_end_bce,
  "MH",          2000,             1600,
  "LH I",        1600,             1500,
  "LH II",       1500,             1390,
  "LH IIIA",     1390,             1300,
  "LH IIIB",     1300,             1190,
  "LH IIIC",     1190,             1070,
  "SM",          1070,             1015
) %>% mutate(
  phase_mid_bce = (phase_start_bce + phase_end_bce) / 2,
  phase_duration = phase_start_bce - phase_end_bce,
  broad_phase = factor(broad_phase,
    levels = c("MH", "LH I", "LH II", "LH IIIA", "LH IIIB", "LH IIIC", "SM"),
    ordered = TRUE)
)

# Assign each grave to broadest phase based on its earliest component
cups_skulls_clean <- cups_skulls_clean %>%
  mutate(
    broad_phase = case_when(
      earliest_component %in% c("MH", "MHII", "MHIII", "MHIIIB") ~ "MH",
      earliest_component %in% c("LH", "LHI")                      ~ "LH I",
      earliest_component %in% c("LHII", "LHIIA", "LHIIB")         ~ "LH II",
      earliest_component %in% c("LHIII", "LHIIIA", "LHIIIA1", "LHIIIA2") ~ "LH IIIA",
      earliest_component %in% c("LHIIIB", "LHIIIB1")              ~ "LH IIIB",
      earliest_component %in% c("LHIIIC", "LHIIIC1")              ~ "LH IIIC",
      earliest_component == "submycenaean"                          ~ "SM",
      TRUE ~ NA_character_
    ),
    broad_phase = factor(broad_phase,
      levels = c("MH", "LH I", "LH II", "LH IIIA", "LH IIIB", "LH IIIC", "SM"),
      ordered = TRUE)
  )

# Map grave_type_simple names to tomb_type dictionary names
grave_to_tomb_map <- tibble(
  grave_type_simple = c("Tholos", "Chamber Tomb", "Cist Grave", "Pit Grave", "Dromos Only", "Other"),
  tomb_type_dict = c("tholos", "chamber tomb", "cist", "pit", "dromos of tholos", NA)
)

# Join tomb type dictionary via mapping
cups_skulls_clean <- cups_skulls_clean %>%
  left_join(grave_to_tomb_map, by = "grave_type_simple") %>%
  left_join(
    tomb_dict %>% select(tomb_type, tomb_type_order),
    by = c("tomb_type_dict" = "tomb_type")
  ) %>%
  select(-tomb_type_dict) %>%

  # Convert to ordered factors
  mutate(
    tomb_type_ordered = factor(
      grave_type_simple,
      levels = grave_to_tomb_map$grave_type_simple[
        order(tomb_dict$tomb_type_order[match(grave_to_tomb_map$tomb_type_dict, tomb_dict$tomb_type)])
      ],
      ordered = TRUE
    ),
    period_ordered = factor(
      period,
      levels = arrange(period_dict, chronological_order) %>% pull(period_code),
      ordered = TRUE
    )
  )

cat("✓ Dictionaries successfully loaded and applied!\n")
cat("New columns created:\n")
cat("  - period_start_order: chronological order of earliest period component\n")
cat("  - period_end_order: chronological order of latest period component\n")
cat("  - period_midpoint_order: midpoint between start and end\n")
cat("  - tomb_type_ordered: ordered factor for tomb types\n")
cat("  - period_ordered: ordered factor for periods\n")
```

## Data Overview

### Basic Statistics

The table below summarizes overall dataset coverage and identifies variables with missing observations. Missing data reflect the state of the archaeological record and publication — not all excavated tombs have complete information on all variables — and are excluded from any calculation where the relevant variable is unknown.

```{r summary-stats}
cat("=== Dataset Summary ===\n\n")
cat("Total graves:", nrow(cups_skulls_clean), "\n")
cat("Graves with drinking vessels recorded:", sum(cups_skulls_clean$has_drinking_vessel == TRUE, na.rm = TRUE), "\n")
cat("Graves with human remains:", sum(cups_skulls_clean$has_remains == TRUE, na.rm = TRUE), "\n")
cat("Graves with hair pins:", sum(cups_skulls_clean$has_pin == TRUE, na.rm = TRUE), "\n")
cat("\n")

# Missing data summary
cat("=== Missing Data ===\n")
cups_skulls_clean %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  dplyr::filter(missing_count > 0) %>%
  arrange(desc(missing_count)) %>%
  kable()
```

### Sites Represented

```{r sites-overview}
sites_summary <- cups_skulls_clean %>%
  count(site, region, name = "n_graves") %>%
  arrange(desc(n_graves))

cat("Number of sites:", n_distinct(cups_skulls_clean$site), "\n\n")
kable(sites_summary,
      col.names = c("Site", "Region", "Number of Graves"),
      caption = "Sites Represented in the Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Grave Type Distribution

Tomb types are compared across regions to establish the structural composition of each sample. Because tholos tombs, chamber tombs, cist graves, and pit graves represent different investments of labor and architectural elaboration, regional differences in tomb type distribution provide context for interpreting other patterns in the assemblage data.

```{r grave-types}
grave_type_summary <- cups_skulls_clean %>%
  count(region, grave_type_simple, name = "count") %>%
  arrange(region, desc(count))

kable(grave_type_summary,
      col.names = c("Region", "Grave Type", "Count"),
      caption = "Grave Type Distribution by Region") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r grave-type-plot, fig.width=12, fig.height=6}
grave_type_summary <- grave_type_summary %>%
  group_by(region) %>%
  mutate(pct = round(count / sum(count) * 100, 1)) %>%
  ungroup()

ggplot(grave_type_summary, aes(x = reorder(grave_type_simple, -count), y = count, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(count, "\n(", pct, "%)")), position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Distribution of Grave Types by Region",
    x = "Grave Type",
    y = "Number of Graves",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Burial Types

### Primary vs Secondary Burials

*Primary burial* refers to the original, articulated interment of a body. *Secondary burial* refers to the later re-deposition of skeletal material — typically disarticulated bones that have been moved, often to make space for a subsequent interment. A single tomb can contain evidence for both. The distribution of these burial categories across regions is examined here as a baseline for later analyses that test whether drinking vessels and hair pins associate preferentially with one burial mode over another.

```{r burial-type-summary}
burial_type_summary <- cups_skulls_clean %>%
  summarise(
    total = n(),
    with_primary = sum(primary_burial == TRUE, na.rm = TRUE),
    with_secondary = sum(secondary_burial == TRUE, na.rm = TRUE),
    both = sum(primary_burial == TRUE & secondary_burial == TRUE, na.rm = TRUE),
    neither = sum(primary_burial == FALSE & secondary_burial == FALSE, na.rm = TRUE),
    unknown = sum(is.na(primary_burial) | is.na(secondary_burial))
  ) %>%
  mutate(
    pct_primary = round(with_primary / total * 100, 1),
    pct_secondary = round(with_secondary / total * 100, 1),
    pct_both = round(both / total * 100, 1)
  )

kable(burial_type_summary,
      col.names = c("Total", "With Primary", "With Secondary", "Both", "Neither", "Unknown",
                    "% Primary", "% Secondary", "% Both"),
      caption = "Primary vs Secondary Burial Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r burial-plot, fig.width=12, fig.height=6}
burial_category_data_region <- cups_skulls_clean %>%
  count(region, burial_category) %>%
  group_by(region) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  ungroup() %>%
  arrange(region, desc(n))

ggplot(burial_category_data_region, aes(x = reorder(burial_category, -n), y = n, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(n, "\n(", pct, "%)")), position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Distribution of Burial Types by Region",
    x = "Burial Type",
    y = "Count",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Drinking Vessels Analysis

Drinking vessels — cups, kylikes, and related forms — are among the most commonly recorded ceramic finds in Mycenaean funerary contexts and have been interpreted as indicators of feasting, libation, or other mortuary ritual. This section documents the prevalence of drinking vessels across the dataset, disaggregates that prevalence by tomb type and burial category, and compares patterns between Laconia and Messenia.

### Overall Drinking Vessel Presence

```{r drinking-vessel-summary}
dv_summary <- cups_skulls_clean %>%
  summarise(
    total_graves = n(),
    with_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    without_vessels = sum(has_drinking_vessel == FALSE, na.rm = TRUE),
    unknown_vessel_status = sum(is.na(has_drinking_vessel))
  ) %>%
  mutate(
    pct_with = round(with_vessels / total_graves * 100, 1),
    avg_vessels_recorded = round(mean(cups_skulls_clean$drinking_vessel_number, na.rm = TRUE), 2)
  )

kable(dv_summary,
      col.names = c("Total Graves", "With Vessels", "Without Vessels", "Unknown", "% With Vessels", "Avg Vessel Count"),
      caption = "Drinking Vessel Presence Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Drinking Vessels by Grave Type

```{r dv-by-grave-type}
dv_by_type <- cups_skulls_clean %>%
  group_by(grave_type_simple) %>%
  summarise(
    total = n(),
    with_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    pct_with_vessels = round(with_vessels / total * 100, 1),
    avg_vessel_count = round(mean(drinking_vessel_number, na.rm = TRUE), 2)
  ) %>%
  arrange(desc(with_vessels))

kable(dv_by_type,
      col.names = c("Grave Type", "Total Graves", "With Vessels", "% With Vessels", "Avg Vessel Count"),
      caption = "Drinking Vessels by Grave Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r dv-by-type-plot, fig.width=12, fig.height=6}
dv_by_type_region <- cups_skulls_clean %>%
  group_by(region, grave_type_simple) %>%
  summarise(
    total = n(),
    with_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    pct_with_vessels = round(with_vessels / total * 100, 1),
    avg_vessel_count = round(mean(drinking_vessel_number, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(region, desc(with_vessels))

ggplot(dv_by_type_region, aes(x = reorder(grave_type_simple, -with_vessels), y = with_vessels, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(with_vessels, " (", pct_with_vessels, "%)")),
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Graves with Drinking Vessels by Grave Type and Region",
    x = "Grave Type",
    y = "Number of Graves with Drinking Vessels",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

### Drinking Vessels by Burial Type

```{r dv-by-burial-type}
dv_by_burial <- cups_skulls_clean %>%
  group_by(burial_category) %>%
  summarise(
    total = n(),
    with_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    pct_with_vessels = round(with_vessels / total * 100, 1),
    avg_vessel_count = round(mean(drinking_vessel_number, na.rm = TRUE), 2)
  ) %>%
  arrange(desc(with_vessels))

kable(dv_by_burial,
      col.names = c("Burial Type", "Total Graves", "With Vessels", "% With Vessels", "Avg Vessel Count"),
      caption = "Drinking Vessels by Burial Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r dv-by-burial-plot, fig.width=12, fig.height=6}
dv_by_burial_region <- cups_skulls_clean %>%
  group_by(region, burial_category) %>%
  summarise(
    total = n(),
    with_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    pct_with_vessels = round(with_vessels / total * 100, 1),
    avg_vessel_count = round(mean(drinking_vessel_number, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(region, desc(with_vessels))

ggplot(dv_by_burial_region, aes(x = reorder(burial_category, -with_vessels), y = with_vessels, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(with_vessels, " (", pct_with_vessels, "%)")),
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Graves with Drinking Vessels by Burial Type and Region",
    x = "Burial Type",
    y = "Number of Graves with Drinking Vessels",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Hair Pin Analysis

Hair pins are personal ornaments found in a small subset of Laconian and Messenian graves. Their presence has been associated with female burials and with certain phases of the Bronze Age, though the sample is too small to support strong generalizations. This section documents overall hair pin prevalence and its regional distribution as a basis for the subsequent analyses examining co-occurrence with secondary burial, drinking vessels, and tomb size.

### Overall Hair Pin Presence

```{r pin-summary}
pin_summary <- cups_skulls_clean %>%
  summarise(
    total_graves = n(),
    with_pins = sum(has_pin == TRUE, na.rm = TRUE),
    without_pins = sum(has_pin == FALSE, na.rm = TRUE),
    unknown_pin_status = sum(is.na(has_pin))
  ) %>%
  mutate(pct_with = round(with_pins / total_graves * 100, 1))

kable(pin_summary,
      col.names = c("Total Graves", "With Pins", "Without Pins", "Unknown", "% With Pins"),
      caption = "Hair Pin Presence Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Hair Pins by Region

```{r pins-by-region, fig.width=12, fig.height=6}
pin_by_region <- cups_skulls_clean %>%
  group_by(region) %>%
  summarise(
    total = n(),
    with_pins = sum(has_pin == TRUE, na.rm = TRUE),
    pct_with_pins = round(with_pins / total * 100, 1),
    .groups = "drop"
  )

ggplot(pin_by_region, aes(x = region, y = with_pins, fill = region)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(with_pins, " (", pct_with_pins, "%)")), vjust = -0.3, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Graves with Hair Pins by Region",
    x = "Region",
    y = "Number of Graves with Hair Pins",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## Secondary Burial and Hair Pins

### Co-occurrence by Region

This section tests whether hair pins are more commonly found in graves with evidence of secondary burial. If pins are associated with specific individuals whose remains were later moved or disturbed, they might be expected to cluster with secondary contexts. Conversely, if pins are associated with the initial primary interment, they might appear more often in graves without secondary activity.

Because the sample of pin-bearing graves is small (n = 16 total), conventional chi-square tests would violate the assumption of adequate cell counts. Fisher's exact test is used instead, as it is appropriate for contingency tables with small expected frequencies. Results are reported separately for each region.

```{r secondary-pins-table}
secondary_pins <- cups_skulls_clean %>%
  dplyr::filter(!is.na(secondary_burial) & !is.na(has_pin)) %>%
  group_by(region, secondary_burial, has_pin) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    secondary_label = ifelse(secondary_burial, "Secondary Present", "Secondary Absent"),
    pin_label = ifelse(has_pin, "With Pin", "No Pin")
  )

# Summary table
secondary_pins_wide <- secondary_pins %>%
  select(region, secondary_label, pin_label, count) %>%
  pivot_wider(names_from = pin_label, values_from = count, values_fill = 0)

kable(secondary_pins_wide)
```

```{r secondary-pins-plot, fig.width=12, fig.height=6}
secondary_pins_pct <- cups_skulls_clean %>%
  dplyr::filter(!is.na(secondary_burial) & !is.na(has_pin)) %>%
  group_by(region, secondary_burial) %>%
  summarise(
    total = n(),
    with_pin = sum(has_pin == TRUE),
    pct_with_pin = round(with_pin / total * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(secondary_label = ifelse(secondary_burial, "Secondary Present", "Secondary Absent"))

ggplot(secondary_pins_pct, aes(x = secondary_label, y = pct_with_pin, fill = region)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  geom_text(aes(label = paste0(with_pin, "/", total, "\n(", pct_with_pin, "%)")),
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Hair Pin Presence by Secondary Burial Status and Region",
    x = "Secondary Burial Status",
    y = "% of Graves with Hair Pins",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

```{r secondary-pins-fisher}
# Fisher's exact test by region (appropriate for small cell counts)
for (reg in unique(cups_skulls_clean$region)) {
  region_data <- cups_skulls_clean %>%
    dplyr::filter(region == reg & !is.na(secondary_burial) & !is.na(has_pin))

  contingency <- table(region_data$secondary_burial, region_data$has_pin)

  cat("\n===", toupper(reg), "===\n")
  print(kable(as.data.frame.matrix(contingency),
              caption = paste("Contingency Table:", toupper(reg))) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

  if (all(dim(contingency) == c(2, 2))) {
    fisher_result <- fisher.test(contingency)
    cat("Fisher's exact test p-value:", round(fisher_result$p.value, 4), "\n")
    cat("Odds ratio:", round(fisher_result$estimate, 3), "\n")
  } else {
    cat("Insufficient variation for statistical test\n")
  }
}
```

## Drinking Vessels, Secondary Burial, and Hair Pins

### Three-Way Cross-Tabulation by Region

To examine whether drinking vessels, secondary burial, and hair pins co-occur beyond chance, a three-way cross-tabulation is constructed separately for each region. This allows identification of any combinations of these three variables that appear disproportionately often or rarely. Given the small total number of pin-bearing graves, this analysis is primarily descriptive; cell counts are reported alongside the visualization.

```{r three-way-table}
three_way <- cups_skulls_clean %>%
  dplyr::filter(!is.na(has_drinking_vessel) & !is.na(secondary_burial) & !is.na(has_pin)) %>%
  group_by(region, has_drinking_vessel, secondary_burial, has_pin) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    vessel_label = ifelse(has_drinking_vessel, "Vessel Present", "No Vessel"),
    secondary_label = ifelse(secondary_burial, "Secondary Present", "Secondary Absent"),
    pin_label = ifelse(has_pin, "With Pin", "No Pin")
  )

kable(three_way %>% select(region, vessel_label, secondary_label, pin_label, count))
```

```{r three-way-plot, fig.width=14, fig.height=8}
three_way_summary <- cups_skulls_clean %>%
  dplyr::filter(!is.na(has_drinking_vessel) & !is.na(secondary_burial) & !is.na(has_pin)) %>%
  mutate(
    vessel_label = ifelse(has_drinking_vessel, "Vessel Present", "No Vessel"),
    secondary_label = ifelse(secondary_burial, "Secondary Present", "Secondary Absent"),
    pin_label = ifelse(has_pin, "With Pin", "No Pin")
  ) %>%
  count(region, vessel_label, secondary_label, pin_label)

ggplot(three_way_summary, aes(x = secondary_label, y = n, fill = pin_label)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("With Pin" = colors$pin, "No Pin" = "grey70")) +
  facet_grid(vessel_label ~ region) +
  geom_text(aes(label = n), position = position_dodge(width = 0.7), vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Hair Pins by Secondary Burial Status and Drinking Vessel Presence",
    subtitle = "Rows: Drinking Vessel status | Columns: Region",
    x = "Secondary Burial Status",
    y = "Count",
    fill = "Hair Pin"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Hair Pins and Chamber Size

Larger tomb chambers may reflect greater investment in funerary architecture, which could correlate with higher-status or more elaborately furnished burials. This section examines whether graves containing hair pins have systematically larger estimated chamber areas than those without. Chamber area is available for a subset of graves and is expressed in estimated square meters.

Because chamber area distributions are right-skewed and may not be normally distributed, a non-parametric Wilcoxon rank-sum test is used to compare the distributions of chamber area between pin-bearing and non-pin-bearing graves within each region. Boxplots are provided to visualize the distributions directly.

### Chamber Area by Hair Pin Presence

```{r pins-area-summary}
pins_area_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(area_of_chamber_estimated) & !is.na(has_pin))

cat("Graves with area AND pin status:", nrow(pins_area_data), "\n")

pins_area_data %>%
  group_by(has_pin) %>%
  summarise(
    n = n(),
    mean_area = round(mean(area_of_chamber_estimated, na.rm = TRUE), 2),
    median_area = round(median(area_of_chamber_estimated, na.rm = TRUE), 2),
    sd_area = round(sd(area_of_chamber_estimated, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  mutate(pin_label = ifelse(has_pin, "With Pin", "No Pin")) %>%
  kable()
```

```{r pins-area-boxplot, fig.width=12, fig.height=6}
ggplot(pins_area_data, aes(x = ifelse(has_pin, "With Pin", "No Pin"),
                            y = area_of_chamber_estimated, fill = region)) +
  geom_boxplot(alpha = 0.7, position = "dodge") +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  labs(
    title = "Chamber Area by Hair Pin Presence and Region",
    x = "Hair Pin Status",
    y = "Chamber Area Estimated (m\u00b2)",
    fill = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

```{r pins-area-test}
# Wilcoxon test (non-parametric, appropriate for potentially non-normal area data)
for (reg in unique(pins_area_data$region)) {
  region_data <- pins_area_data %>% dplyr::filter(region == reg)
  with_pin <- region_data %>% dplyr::filter(has_pin == TRUE) %>% pull(area_of_chamber_estimated)
  without_pin <- region_data %>% dplyr::filter(has_pin == FALSE) %>% pull(area_of_chamber_estimated)

  cat("\n===", toupper(reg), "===\n")
  cat("With pin: n =", length(with_pin), ", median =", round(median(with_pin), 2), "\n")
  cat("Without pin: n =", length(without_pin), ", median =", round(median(without_pin), 2), "\n")

  if (length(with_pin) >= 2 & length(without_pin) >= 2) {
    test_result <- wilcox.test(with_pin, without_pin)
    cat("Wilcoxon rank-sum test p-value:", round(test_result$p.value, 4), "\n")
  } else {
    cat("Too few observations for statistical test\n")
  }
}
```

## Skeletal Remains Analysis

The presence of identifiable human remains — particularly skulls, which are the most commonly recorded skeletal element — provides a partial window into how many individuals were interred and whether bones were disturbed or removed. This section describes overall remains prevalence and skull counts, which are used in subsequent analyses of the relationship between drinking vessels and the physical presence of the dead.

### Remains Distribution

```{r remains-summary}
remains_summary <- cups_skulls_clean %>%
  summarise(
    total_graves = n(),
    with_remains = sum(has_remains == TRUE, na.rm = TRUE),
    without_remains = sum(has_remains == FALSE, na.rm = TRUE),
    unknown_remains = sum(is.na(has_remains))
  ) %>%
  mutate(pct_with_remains = round(with_remains / total_graves * 100, 1))

kable(remains_summary,
      col.names = c("Total Graves", "With Remains", "Without Remains", "Unknown", "% With Remains"),
      caption = "Skeletal Remains Presence Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Skulls Recorded

```{r skulls-analysis}
skulls_stats <- cups_skulls_clean %>%
  dplyr::filter(!is.na(skulls_number)) %>%
  summarise(
    graves_with_skull_count = n(),
    total_skulls = sum(skulls_number, na.rm = TRUE),
    avg_skulls = round(mean(skulls_number, na.rm = TRUE), 2),
    min_skulls = min(skulls_number, na.rm = TRUE),
    max_skulls = max(skulls_number, na.rm = TRUE)
  )

kable(skulls_stats,
      col.names = c("Graves with Skull Count", "Total Skulls", "Avg Skulls", "Min", "Max"),
      caption = "Skull Count Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r skulls-distribution}
skulls_by_count <- cups_skulls_clean %>%
  dplyr::filter(!is.na(skulls_number)) %>%
  count(skulls_number, name = "graves") %>%
  arrange(skulls_number)

kable(skulls_by_count,
      col.names = c("Skulls Recorded", "Number of Graves"),
      caption = "Distribution of Skull Counts per Grave") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Drinking Vessels and Skeletal Remains

A central question for this thesis is whether drinking vessels are more likely to be deposited when human remains are present in the tomb — suggesting a direct relationship between the objects and the body — or whether they appear regardless of skeletal preservation. This section cross-tabulates drinking vessel presence against skeletal remains presence, and examines whether the number of drinking vessels correlates with the number of skulls recorded (as a proxy for the number of individuals interred).

### Drinking Vessel Presence and Remains Co-occurrence

```{r vessels-and-remains}
vessels_remains_cross <- cups_skulls_clean %>%
  dplyr::filter(!is.na(has_drinking_vessel) & !is.na(has_remains)) %>%
  count(has_drinking_vessel, has_remains) %>%
  pivot_wider(names_from = has_remains, values_from = n, values_fill = 0) %>%
  rename(no_remains = `FALSE`, with_remains = `TRUE`)

vessels_remains_cross <- vessels_remains_cross %>%
  mutate(vessel_status = ifelse(has_drinking_vessel, "With Drinking Vessels", "No Drinking Vessels")) %>%
  select(vessel_status, no_remains, with_remains)

kable(vessels_remains_cross,
      col.names = c("Vessel Status", "No Remains", "With Remains"),
      caption = "Cross-Tabulation: Drinking Vessels and Skeletal Remains") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r vessels-and-remains-analysis}
# Calculate cross-tabulation with percentages
cross_analysis <- cups_skulls_clean %>%
  dplyr::filter(!is.na(has_drinking_vessel) & !is.na(has_remains)) %>%
  group_by(has_drinking_vessel) %>%
  summarise(
    total = n(),
    with_remains = sum(has_remains == TRUE, na.rm = TRUE),
    pct_with_remains = round(with_remains / total * 100, 1)
  ) %>%
  mutate(vessel_status = ifelse(has_drinking_vessel, "With Drinking Vessels", "Without Drinking Vessels"))

kable(select(cross_analysis, vessel_status, total, with_remains, pct_with_remains))
```

### Drinking Vessels and Skull Count

```{r vessels-skulls-scatter}
vessels_skulls_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(drinking_vessel_number) & !is.na(skulls_number)) %>%
  mutate(has_vessels_label = ifelse(has_drinking_vessel, "With Vessels", "No Vessels"))

ggplot(vessels_skulls_data, aes(x = drinking_vessel_number, y = skulls_number, color = has_vessels_label)) +
  geom_jitter(size = 3, alpha = 0.6, width = 0.2, height = 0.2) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.2) +
  labs(
    title = "Relationship: Drinking Vessels vs Skull Count",
    x = "Number of Drinking Vessels Recorded",
    y = "Number of Skulls",
    color = ""
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## Regional Comparison

This section brings together the key assemblage variables into a single regional summary, directly comparing Laconia and Messenia on drinking vessel prevalence, skeletal remains presence, and average skull count per grave. Differences between regions are relevant for interpreting whether patterns observed in the full dataset are driven primarily by one region or reflect a broader Mycenaean phenomenon.

### Summary by Region

```{r regional-summary}
regional_summary <- cups_skulls_clean %>%
  group_by(region) %>%
  summarise(
    total_graves = n(),
    with_drinking_vessels = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    pct_with_vessels = round(with_drinking_vessels / total_graves * 100, 1),
    with_remains = sum(has_remains == TRUE, na.rm = TRUE),
    pct_with_remains = round(with_remains / total_graves * 100, 1),
    avg_skulls = round(mean(skulls_number, na.rm = TRUE), 2)
  )

kable(regional_summary,
      col.names = c("Region", "Total Graves", "With Drinking Vessels", "% With Vessels",
                    "With Remains", "% With Remains", "Avg Skulls"),
      caption = "Regional Summary: Drinking Vessels and Skeletal Remains") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Chronological Overview

### Periods Represented in the Dataset

Graves in this dataset carry period attributions that range in precision from a single sub-phase (e.g., "LHIIIA2") to broad spans covering several centuries (e.g., "MH–LHIIIC"). The table below summarizes how many graves fall into broad period categories. Graves with no recorded period are retained in all non-temporal analyses but excluded from analyses that require chronological assignment.

```{r period-distribution}
period_summary <- cups_skulls_clean %>%
  mutate(
    period_category = case_when(
      str_detect(tolower(period), "mh") ~ "Middle Helladic",
      str_detect(tolower(period), "lh\\s*i(?!i)")  | str_detect(tolower(period), "lh i-") ~ "LH I-II",
      str_detect(tolower(period), "lh\\s*ii") & !str_detect(tolower(period), "lh\\s*ii") ~ "LH II",
      str_detect(tolower(period), "lh\\s*iii") ~ "LH III",
      str_detect(tolower(period), "lh") ~ "Late Helladic (general)",
      is.na(period) | period == "" ~ "Unrecorded",
      TRUE ~ "Other"
    )
  ) %>%
  count(period_category, name = "graves") %>%
  arrange(desc(graves))

kable(period_summary,
      col.names = c("Period Category", "Number of Graves"),
      caption = "Graves by Period Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Period Span Analysis

The chart below maps each period code used in the dataset onto the Bronze Age calendar. The x-axis is proportional to actual calendar years (BCE), making the width of each row a direct representation of chronological uncertainty. Range periods (e.g., "LHIIA–LHIIIB") appear as horizontal segments; single-phase codes (e.g., "LHIIIA2") appear as points. Dashed vertical lines mark the start of each major phase; tick labels that combine multiple codes (e.g., "LHIII / LHIIIA / LHIIIA1") reflect phases whose calendar start dates coincide.

This visualization is intended to make the degree of chronological imprecision in the dataset transparent before the temporal analyses that follow.

```{r period-span-chart, fig.width=16, fig.height=14}
# Identify single-component periods and attach calendar years
single_periods <- period_dict %>%
  dplyr::filter(earliest_component == latest_component) %>%
  select(period_code, chronological_order) %>%
  arrange(chronological_order) %>%
  distinct(period_code, .keep_all = TRUE) %>%
  dplyr::filter(period_code != "LH II") %>%
  left_join(component_cal %>% select(component, cal_start_bce),
            by = c("period_code" = "component"))

# Get all period codes in the data with their calendar year spans
period_spans <- cups_skulls_clean %>%
  dplyr::filter(!is.na(period_start_cal) & !is.na(period_end_cal)) %>%
  distinct(period, period_start_cal, period_end_cal) %>%
  mutate(
    is_range = period_start_cal != period_end_cal,
    span_width = period_start_cal - period_end_cal  # BCE: larger start = earlier
  ) %>%
  arrange(desc(period_start_cal), period_end_cal)

# Count graves per period code
period_counts <- cups_skulls_clean %>%
  dplyr::filter(!is.na(period_start_cal)) %>%
  count(period, name = "n_graves")

period_spans <- period_spans %>%
  left_join(period_counts, by = "period") %>%
  mutate(period_label = paste0(period, " (n=", n_graves, ")")) %>%
  mutate(period_label = factor(period_label,
    levels = rev(mutate(arrange(., desc(period_start_cal), period_end_cal),
                        period_label = paste0(period, " (n=", n_graves, ")"))$period_label)))

# Collapse period codes that share the same calendar position into one label
# (e.g. LH and LHI both start at 1600 BCE — combine as "LH / LHI")
single_periods_unique <- dplyr::filter(single_periods, !is.na(cal_start_bce)) %>%
  group_by(cal_start_bce) %>%
  summarise(period_code = paste(period_code, collapse = " / "), .groups = "drop")

ggplot(period_spans) +
  geom_segment(
    data = dplyr::filter(period_spans, is_range),
    aes(x = period_start_cal, xend = period_end_cal,
        y = period_label, yend = period_label),
    linewidth = 2, color = colors$drinking_vessel, alpha = 0.7
  ) +
  geom_point(
    aes(x = period_start_cal, y = period_label),
    size = 3, color = colors$drinking_vessel
  ) +
  geom_point(
    data = dplyr::filter(period_spans, is_range),
    aes(x = period_end_cal, y = period_label),
    size = 3, color = colors$drinking_vessel
  ) +
  geom_vline(
    data = single_periods_unique,
    aes(xintercept = cal_start_bce),
    linetype = "dashed", alpha = 0.3
  ) +
  scale_x_reverse(
    breaks = single_periods_unique$cal_start_bce,
    labels = single_periods_unique$period_code
  ) +
  labs(
    title = "Period Spans in the Dataset",
    subtitle = "Each row is a period code used in the data. X-axis is proportional calendar time (BCE).",
    x = "Calendar Date (BCE)",
    y = "Period Code (n = number of graves)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank()
  )
```

```{r period-span-summary}
cat("=== Individual chronological components (in order) ===\n\n")
for (i in 1:nrow(single_periods)) {
  cat(sprintf("  %2d. %s\n", i, single_periods$period_code[i]))
}

cat("\n=== Summary ===\n")
cat("Single-period codes:", sum(!period_spans$is_range), "\n")
cat("Range-period codes:", sum(period_spans$is_range), "\n")
cat("Total period codes in data:", nrow(period_spans), "\n")

cat("\n=== Range periods spanning 3+ components ===\n")
wide_spans <- period_spans %>%
  dplyr::filter(span_width > 10) %>%
  arrange(desc(span_width))
if (nrow(wide_spans) > 0) {
  for (i in 1:nrow(wide_spans)) {
    cat(sprintf("  %s: spans from order %g to %g (width = %g, n = %d graves)\n",
        wide_spans$period[i], wide_spans$period_start_order[i],
        wide_spans$period_end_order[i], wide_spans$span_width[i],
        wide_spans$n_graves[i]))
  }
}
```

### Individual Graves by Period Span and Region

Each line represents a single grave plotted against calendar time, spanning its full attributed date range. Graves are colored by hair pin presence, allowing visual inspection of whether pin-bearing graves are concentrated in particular periods or distributed broadly across the sequence. Graves without datable period codes are excluded.

```{r graves-by-period-region, fig.width=16, fig.height=14}
# Prepare individual grave data with calendar year spans
graves_time <- cups_skulls_clean %>%
  dplyr::filter(!is.na(period_start_cal) & !is.na(period_end_cal) & !is.na(has_pin)) %>%
  mutate(pin_label = ifelse(has_pin, "With Pin", "No Pin")) %>%
  # Order graves: earlier start first (higher BCE = earlier), then narrower range
  arrange(region, desc(period_start_cal), period_end_cal) %>%
  group_by(region) %>%
  mutate(grave_rank = row_number()) %>%
  ungroup()

# Reuse the deduplicated tick table built in the period-span-chart chunk
# (collapses codes sharing the same calendar position, e.g. "LHIII / LHIIIA / LHIIIA1")

ggplot(graves_time) +
  geom_segment(
    aes(x = period_start_cal, xend = period_end_cal,
        y = grave_rank, yend = grave_rank,
        color = pin_label),
    linewidth = 0.8, alpha = 0.7
  ) +
  geom_point(
    data = dplyr::filter(graves_time, period_start_cal == period_end_cal),
    aes(x = period_start_cal, y = grave_rank, color = pin_label),
    size = 1.5, alpha = 0.7
  ) +
  scale_color_manual(values = c("With Pin" = colors$pin, "No Pin" = colors$drinking_vessel)) +
  geom_vline(
    data = single_periods_unique,
    aes(xintercept = cal_start_bce),
    linetype = "dashed", alpha = 0.2
  ) +
  scale_x_reverse(
    breaks = single_periods_unique$cal_start_bce,
    labels = single_periods_unique$period_code
  ) +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  labs(
    title = "Individual Graves by Period Span and Region",
    subtitle = "Each line = one grave. X-axis is proportional calendar time (BCE). Colored by hair pin presence.",
    x = "Calendar Date (BCE)",
    y = "Graves (ordered by period)",
    color = "Hair Pin"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

## Temporal Analyses

These analyses examine whether funerary practices — burial type, drinking vessel presence, hair pin presence, and tomb size — changed over the course of the Bronze Age, and whether any such changes differed between Laconia and Messenia.

### Phase Assignment

Because most graves carry date ranges rather than precise phase attributions, each grave is assigned to a broad phase based on the **earliest** component of its recorded period. A grave dated "LHIIA–LHIIIB," for instance, is counted in the LH II phase. This is a conservative choice: it avoids assigning a grave to a later phase it may not actually represent, but it means that graves with wide date ranges are anchored at their earliest possible position. Any apparent trends toward earlier phases should be read with this in mind.

Seven broad phases are used: MH, LH I, LH II, LH III, LH IIIA, LH IIIB, and LH IIIC/Submycenaean. Graves whose period code could not be matched to any known component receive `NA` and are excluded from all temporal plots. Graves recorded only as the undifferentiated "LH" code are assigned to LH I.

Each temporal plot is shown in two versions: **Option A** uses an ordered categorical x-axis with equal spacing between phases; **Option B** uses a proportional calendar x-axis where bar widths are scaled to actual phase duration in calendar years. Option B better represents the unequal lengths of the phases but can be harder to read when phases have few graves.

```{r phase-coverage}
phase_table <- as.data.frame(table(cups_skulls_clean$broad_phase, useNA = "ifany"))
names(phase_table) <- c("Broad Phase", "Number of Graves")
kable(phase_table, caption = "Graves Assigned to Broad Phases") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

### Burial Type Composition Over Time

These charts track whether the proportion of primary-only, secondary-only, mixed, and unrecorded burials changed across the Bronze Age sequence within each region. Stacked bars represent 100% of graves assigned to each phase, so the chart shows relative composition rather than absolute counts. The "Neither/Unknown" category reflects graves for which burial type was not recorded in the published sources; its size in any given phase indicates the limits of the available evidence for that period rather than the absence of burials.

#### Option A: Broad Phase Categories — Proportions

```{r burial-over-time, fig.width=14, fig.height=8}
burial_cat_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(broad_phase)) %>%
  mutate(burial_cat = case_when(
    primary_burial == TRUE & secondary_burial == TRUE ~ "Both",
    primary_burial == TRUE                            ~ "Primary only",
    secondary_burial == TRUE                          ~ "Secondary only",
    TRUE                                              ~ "Neither/Unknown"
  ),
  burial_cat = factor(burial_cat,
    levels = c("Primary only", "Both", "Secondary only", "Neither/Unknown"))) %>%
  group_by(region, broad_phase, burial_cat) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(region, broad_phase) %>%
  mutate(total = sum(n), pct = n / total * 100) %>%
  ungroup()

phase_totals_burial <- burial_cat_data %>%
  distinct(region, broad_phase, total)

ggplot(burial_cat_data, aes(x = broad_phase, y = pct, fill = burial_cat)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = ifelse(pct >= 8, paste0(round(pct), "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  geom_text(data = phase_totals_burial,
            aes(x = broad_phase, y = 104, label = paste0("n=", total)),
            inherit.aes = FALSE, vjust = 0, size = 3.5) +
  scale_fill_manual(values = c(
    "Primary only"    = colors$drinking_vessel,
    "Both"            = "#44AA99",
    "Secondary only"  = colors$skulls,
    "Neither/Unknown" = "grey80"
  )) +
  scale_y_continuous(limits = c(0, 113), labels = function(x) paste0(x, "%")) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Type Composition Over Time by Region",
    subtitle = "Option A: each bar = 100% of graves in that phase; n= total graves",
    x = "Broad Phase", y = "% of Graves in Phase", fill = "Burial Type"
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option A: Raw Counts

```{r burial-over-time-raw, fig.width=14, fig.height=8}
ggplot(burial_cat_data, aes(x = broad_phase, y = n, fill = burial_cat)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = ifelse(n >= 2, n, "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  scale_fill_manual(values = c(
    "Primary only"    = colors$drinking_vessel,
    "Both"            = "#44AA99",
    "Secondary only"  = colors$skulls,
    "Neither/Unknown" = "grey80"
  )) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Type Counts Over Time by Region",
    subtitle = "Option A: raw number of graves per burial type category per phase",
    x = "Broad Phase", y = "Number of Graves", fill = "Burial Type"
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option B: Proportional Calendar Time (BCE) — Proportions

```{r burial-over-time-b, fig.width=14, fig.height=8}
burial_cat_b <- burial_cat_data %>%
  left_join(phase_cal, by = "broad_phase") %>%
  mutate(bar_width = phase_duration * 0.75)

phase_totals_burial_b <- burial_cat_b %>%
  distinct(region, broad_phase, total, phase_mid_bce)

ggplot(burial_cat_b, aes(x = phase_mid_bce, y = pct, fill = burial_cat, width = bar_width)) +
  geom_col(position = "stack") +
  geom_text(aes(label = ifelse(pct >= 8, paste0(round(pct), "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  geom_text(data = phase_totals_burial_b,
            aes(x = phase_mid_bce, y = 104, label = paste0("n=", total)),
            inherit.aes = FALSE, vjust = 0, size = 3.5) +
  scale_fill_manual(values = c(
    "Primary only"    = colors$drinking_vessel,
    "Both"            = "#44AA99",
    "Secondary only"  = colors$skulls,
    "Neither/Unknown" = "grey80"
  )) +
  scale_y_continuous(limits = c(0, 113), labels = function(x) paste0(x, "%")) +
  scale_x_reverse(
    breaks = phase_cal$phase_start_bce,
    labels = paste0(phase_cal$phase_start_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Type Composition Over Time by Region",
    subtitle = "Option B: bar widths proportional to phase duration; each bar = 100% of graves in that phase",
    x = "Calendar Date BCE (phase start)", y = "% of Graves in Phase", fill = "Burial Type"
  ) +
  theme(legend.position = "bottom")
```

#### Option B: Raw Counts

```{r burial-over-time-b-raw, fig.width=14, fig.height=8}
ggplot(burial_cat_b, aes(x = phase_mid_bce, y = n, fill = burial_cat, width = bar_width)) +
  geom_col(position = "stack") +
  geom_text(aes(label = ifelse(n >= 2, n, "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  scale_fill_manual(values = c(
    "Primary only"    = colors$drinking_vessel,
    "Both"            = "#44AA99",
    "Secondary only"  = colors$skulls,
    "Neither/Unknown" = "grey80"
  )) +
  scale_x_reverse(
    breaks = phase_cal$phase_start_bce,
    labels = paste0(phase_cal$phase_start_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Type Counts Over Time by Region",
    subtitle = "Option B: raw counts; bar widths proportional to phase duration",
    x = "Calendar Date BCE (phase start)", y = "Number of Graves", fill = "Burial Type"
  ) +
  theme(legend.position = "bottom")
```

---

### Regional Comparison: Secondary Burial Over Time

These plots isolate the **secondary burial rate** — the proportion of graves in each phase with evidence for secondary burial — and compare Laconia and Messenia directly. The two regions are stacked top-to-bottom to make the x-axis comparable at a glance.

**What to look for:** Do the two regions track together (suggesting pan-regional practice) or diverge (suggesting regional differences in how and when secondary burial became common)? Phases with very small n (shown in the labels) should be interpreted cautiously — a single grave shifts the percentage dramatically.

Labels show "X% (N secondary of M total graves in that phase)."

#### Option A: Broad Phase Categories — Proportions

```{r burial-time-comparison, fig.width=14, fig.height=8}
secondary_phase <- cups_skulls_clean %>%
  dplyr::filter(!is.na(broad_phase) & !is.na(secondary_burial)) %>%
  group_by(region, broad_phase) %>%
  summarise(
    n_graves    = n(),
    n_secondary = sum(secondary_burial == TRUE),
    pct_secondary = round(n_secondary / n_graves * 100, 1),
    .groups = "drop"
  )

ggplot(secondary_phase, aes(x = broad_phase, y = pct_secondary, fill = region)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(pct_secondary), "%\n(", n_secondary, " of ", n_graves, ")")),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  scale_y_continuous(limits = c(0, 130)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Secondary Burial Frequency Over Time: Laconia vs Messenia",
    subtitle = "Option A: % of graves in each phase with secondary burial; label = secondary of total",
    x = "Broad Phase", y = "% of Graves with Secondary Burial", fill = "Region"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option A: Raw Counts

```{r burial-time-comparison-raw, fig.width=14, fig.height=8}
secondary_raw_stack <- secondary_phase %>%
  mutate(n_other = n_graves - n_secondary) %>%
  pivot_longer(cols = c(n_secondary, n_other),
               names_to = "burial_type", values_to = "count") %>%
  mutate(burial_type = ifelse(burial_type == "n_secondary", "Secondary Burial", "Other Graves"))

ggplot(secondary_raw_stack, aes(x = broad_phase, y = count, fill = burial_type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = ifelse(count > 0, count, "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  scale_fill_manual(values = c("Secondary Burial" = colors$skulls, "Other Graves" = "grey80")) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Secondary Burial Counts Over Time: Laconia vs Messenia",
    subtitle = "Option A: each bar = total graves in phase; highlighted segment = graves with secondary burial",
    x = "Broad Phase", y = "Number of Graves", fill = ""
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option B: Proportional Calendar Time (BCE) — Proportions

```{r burial-time-comparison-b, fig.width=14, fig.height=8}
secondary_phase_b <- secondary_phase %>%
  left_join(phase_cal, by = "broad_phase") %>%
  mutate(bar_width = phase_duration * 0.8)

ggplot(secondary_phase_b,
       aes(x = phase_mid_bce, y = pct_secondary, fill = region, width = bar_width)) +
  geom_col(position = "identity") +
  geom_text(aes(label = paste0(round(pct_secondary), "%\n(", n_secondary, " of ", n_graves, ")")),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  scale_y_continuous(limits = c(0, 130)) +
  scale_x_reverse(
    breaks = phase_cal$phase_start_bce,
    labels = paste0(phase_cal$phase_start_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Secondary Burial Frequency Over Time: Laconia vs Messenia",
    subtitle = "Option B: bar widths proportional to phase duration; % of graves in that phase with secondary burial",
    x = "Calendar Date BCE (phase start)", y = "% of Graves with Secondary Burial", fill = "Region"
  ) +
  theme(legend.position = "none")
```

#### Option B: Raw Counts

```{r burial-time-comparison-b-raw, fig.width=14, fig.height=8}
secondary_raw_stack_b <- secondary_phase_b %>%
  mutate(n_other = n_graves - n_secondary) %>%
  pivot_longer(cols = c(n_secondary, n_other),
               names_to = "burial_type", values_to = "count") %>%
  mutate(burial_type = ifelse(burial_type == "n_secondary", "Secondary Burial", "Other Graves"))

ggplot(secondary_raw_stack_b,
       aes(x = phase_mid_bce, y = count, fill = burial_type, width = bar_width)) +
  geom_col(position = "stack") +
  geom_text(aes(label = ifelse(count > 0, count, "")),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  scale_fill_manual(values = c("Secondary Burial" = colors$skulls, "Other Graves" = "grey80")) +
  scale_x_reverse(
    breaks = phase_cal$phase_start_bce,
    labels = paste0(phase_cal$phase_start_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Secondary Burial Counts Over Time: Laconia vs Messenia",
    subtitle = "Option B: raw counts; bar widths proportional to phase duration",
    x = "Calendar Date BCE (phase start)", y = "Number of Graves", fill = ""
  ) +
  theme(legend.position = "bottom")
```

---

### All Burial Variables Over Time

These line charts bring together the three main assemblage attributes — drinking vessels, secondary burial, and hair pins — on a single timeline, allowing comparison of their trends. Point size reflects the total number of graves in that phase and region, so visually smaller points signal phases with sparse data where percentages are less reliable.

**Important:** The three variables are **not mutually exclusive** — a grave may have drinking vessels *and* secondary burial *and* a hair pin. The lines therefore do not add up to 100%. Each line independently answers: "In what share of graves in this phase is this attribute present?"

**Excluded graves:** Graves with `NA` for `broad_phase` (unmatched period codes) are not plotted. This is typically a small number; see the phase coverage table above.

**Interpreting the raw count companion:** The raw count version makes the denominators visible. When a percentage looks high or low, checking the raw counts reveals whether it is based on 2 graves or 20. Phases where total graves (point size) are very small should be treated as preliminary observations, not firm patterns.

#### Option A: Broad Phase Categories — Proportions

```{r vessels-pins-time, fig.width=10, fig.height=8}
multi_phase_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(broad_phase)) %>%
  group_by(region, broad_phase) %>%
  summarise(
    n_graves          = n(),
    n_drinking_vessel = sum(has_drinking_vessel == TRUE, na.rm = TRUE),
    n_secondary       = sum(secondary_burial    == TRUE, na.rm = TRUE),
    n_pin             = sum(has_pin             == TRUE, na.rm = TRUE),
    pct_drinking_vessel = round(n_drinking_vessel / n_graves * 100, 1),
    pct_secondary       = round(n_secondary       / n_graves * 100, 1),
    pct_pin             = round(n_pin             / n_graves * 100, 1),
    .groups = "drop"
  )

multi_phase_pct_long <- multi_phase_data %>%
  pivot_longer(cols = c(pct_drinking_vessel, pct_secondary, pct_pin),
               names_to = "variable", values_to = "percentage") %>%
  mutate(variable = case_when(
    variable == "pct_drinking_vessel" ~ "Drinking Vessel",
    variable == "pct_secondary"       ~ "Secondary Burial",
    variable == "pct_pin"             ~ "Hair Pin"
  ),
  variable = factor(variable, levels = c("Drinking Vessel", "Secondary Burial", "Hair Pin")))

dodge_a <- position_dodge(width = 0.4)

ggplot(multi_phase_pct_long,
       aes(x = broad_phase, y = percentage, color = variable, group = variable)) +
  geom_line(linewidth = 1.2, position = dodge_a) +
  geom_point(aes(size = n_graves), position = dodge_a) +
  geom_text_repel(aes(label = paste0(round(percentage), "%")),
                  size = 4, show.legend = FALSE,
                  bg.color = "white", bg.r = 0.15,
                  min.segment.length = 0, seed = 42,
                  box.padding = 0.4,
                  position = dodge_a) +
  scale_color_manual(values = c(
    "Drinking Vessel"  = colors$drinking_vessel,
    "Secondary Burial" = colors$skulls,
    "Hair Pin"         = colors$pin
  )) +
  scale_size_continuous(range = c(2, 8)) +
  scale_y_continuous(limits = c(0, 115)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Practices Over Time by Region",
    subtitle = "Option A: % of graves with each attribute per phase; point size = total graves\nNote: variables are not mutually exclusive — a grave may appear in multiple lines",
    x = "Broad Phase", y = "% of Graves in Phase",
    color = "Variable", size = "Total Graves"
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option A: Raw Counts

```{r vessels-pins-time-raw, fig.width=10, fig.height=8}
multi_phase_n_long <- multi_phase_data %>%
  pivot_longer(cols = c(n_drinking_vessel, n_secondary, n_pin),
               names_to = "variable", values_to = "count") %>%
  mutate(variable = case_when(
    variable == "n_drinking_vessel" ~ "Drinking Vessel",
    variable == "n_secondary"       ~ "Secondary Burial",
    variable == "n_pin"             ~ "Hair Pin"
  ),
  variable = factor(variable, levels = c("Drinking Vessel", "Secondary Burial", "Hair Pin")))

ggplot(multi_phase_n_long,
       aes(x = broad_phase, y = count, color = variable, group = variable)) +
  geom_line(linewidth = 1.2, position = dodge_a) +
  geom_point(aes(size = n_graves), position = dodge_a) +
  geom_text_repel(aes(label = paste0(count, " of ", n_graves)),
                  size = 4, show.legend = FALSE,
                  bg.color = "white", bg.r = 0.15,
                  min.segment.length = 0, seed = 42,
                  box.padding = 0.4,
                  position = dodge_a) +
  scale_color_manual(values = c(
    "Drinking Vessel"  = colors$drinking_vessel,
    "Secondary Burial" = colors$skulls,
    "Hair Pin"         = colors$pin
  )) +
  scale_size_continuous(range = c(2, 8)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Practice Counts Over Time by Region",
    subtitle = "Option A: raw count with each attribute per phase; label = count of total graves in phase\nNote: variables are not mutually exclusive — a grave may appear in multiple lines",
    x = "Broad Phase", y = "Number of Graves with Attribute",
    color = "Variable", size = "Total Graves"
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option B: Proportional Calendar Time (BCE) — Proportions

```{r vessels-pins-time-b, fig.width=10, fig.height=8}
multi_phase_b_pct <- multi_phase_pct_long %>%
  left_join(phase_cal, by = "broad_phase") %>%
  mutate(x_plot = phase_mid_bce + case_when(
    variable == "Drinking Vessel"  ~ -8L,
    variable == "Secondary Burial" ~  0L,
    variable == "Hair Pin"         ~  8L
  ))

ggplot(multi_phase_b_pct,
       aes(x = x_plot, y = percentage, color = variable, group = variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(size = n_graves)) +
  geom_text_repel(aes(label = paste0(round(percentage), "%")),
                  size = 4, show.legend = FALSE,
                  bg.color = "white", bg.r = 0.15,
                  min.segment.length = 0, seed = 42,
                  box.padding = 0.4) +
  scale_color_manual(values = c(
    "Drinking Vessel"  = colors$drinking_vessel,
    "Secondary Burial" = colors$skulls,
    "Hair Pin"         = colors$pin
  )) +
  scale_size_continuous(range = c(2, 8)) +
  scale_y_continuous(limits = c(0, 115)) +
  scale_x_reverse(
    breaks = phase_cal$phase_mid_bce,
    labels = paste0(phase_cal$phase_mid_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Practices Over Time by Region",
    subtitle = "Option B: proportional calendar time; point size = total graves in phase; points offset ±8 yr for clarity\nNote: variables are not mutually exclusive — a grave may appear in multiple lines",
    x = "Calendar Date BCE (phase midpoint)", y = "% of Graves in Phase",
    color = "Variable", size = "Total Graves"
  ) +
  theme(legend.position = "bottom")
```

#### Option B: Raw Counts

```{r vessels-pins-time-b-raw, fig.width=10, fig.height=8}
multi_phase_b_n <- multi_phase_n_long %>%
  left_join(phase_cal, by = "broad_phase") %>%
  mutate(x_plot = phase_mid_bce + case_when(
    variable == "Drinking Vessel"  ~ -8L,
    variable == "Secondary Burial" ~  0L,
    variable == "Hair Pin"         ~  8L
  ))

ggplot(multi_phase_b_n,
       aes(x = x_plot, y = count, color = variable, group = variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(size = n_graves)) +
  geom_text_repel(aes(label = paste0(count, " of ", n_graves)),
                  size = 4, show.legend = FALSE,
                  bg.color = "white", bg.r = 0.15,
                  min.segment.length = 0, seed = 42,
                  box.padding = 0.4) +
  scale_color_manual(values = c(
    "Drinking Vessel"  = colors$drinking_vessel,
    "Secondary Burial" = colors$skulls,
    "Hair Pin"         = colors$pin
  )) +
  scale_size_continuous(range = c(2, 8)) +
  scale_x_reverse(
    breaks = phase_cal$phase_mid_bce,
    labels = paste0(phase_cal$phase_mid_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Burial Practice Counts Over Time by Region",
    subtitle = "Option B: raw counts; proportional calendar time; points offset ±8 yr for clarity\nNote: variables are not mutually exclusive — a grave may appear in multiple lines",
    x = "Calendar Date BCE (phase midpoint)", y = "Number of Graves with Attribute",
    color = "Variable", size = "Total Graves"
  ) +
  theme(legend.position = "bottom")
```

---

### Chamber Area Over Time

These boxplots examine whether estimated tomb chamber size changed across the Bronze Age sequence. Chamber area is expressed in square meters and is available for a subset of graves; sample sizes per phase are noted above each box. Phases with three or fewer graves produce unreliable box statistics and should be interpreted cautiously. Because tholos tombs are structurally larger than cist or pit graves, any apparent temporal trend in chamber size should be read alongside the tomb type distribution shown earlier.

#### Option A: Broad Phase Categories

```{r area-over-time, fig.width=14, fig.height=7}
area_phase_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(broad_phase) & !is.na(area_of_chamber_estimated))

area_n <- area_phase_data %>%
  group_by(region, broad_phase) %>%
  summarise(n = n(), .groups = "drop")

ggplot(area_phase_data, aes(x = broad_phase, y = area_of_chamber_estimated, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  geom_text(data = area_n,
            aes(x = broad_phase, y = Inf, label = paste0("n=", n)),
            inherit.aes = FALSE, vjust = 2, size = 3.5) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Chamber Area Over Time by Region",
    subtitle = "Option A: boxplots per broad phase; n= graves with area data in that phase",
    x = "Broad Phase", y = "Chamber Area Estimated (m\u00b2)", fill = "Region"
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Option B: Proportional Calendar Time (BCE)

```{r area-over-time-b, fig.width=14, fig.height=7}
area_phase_b <- area_phase_data %>%
  left_join(phase_cal, by = "broad_phase")

area_n_b <- area_phase_b %>%
  group_by(region, broad_phase, phase_mid_bce) %>%
  summarise(n = n(), .groups = "drop")

ggplot(area_phase_b, aes(x = phase_mid_bce, y = area_of_chamber_estimated,
                          group = interaction(broad_phase, region), fill = region)) +
  geom_boxplot(aes(width = phase_duration * 0.8), alpha = 0.7) +
  geom_text(data = area_n_b,
            aes(x = phase_mid_bce, y = Inf, label = paste0("n=", n)),
            inherit.aes = FALSE, vjust = 2, size = 3.5) +
  scale_fill_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  scale_x_reverse(
    breaks = phase_cal$phase_start_bce,
    labels = paste0(phase_cal$phase_start_bce, "\n(", phase_cal$broad_phase, ")")
  ) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Chamber Area Over Time by Region",
    subtitle = "Option B: box widths proportional to phase duration; n= graves with area data",
    x = "Calendar Date BCE (phase start)", y = "Chamber Area Estimated (m\u00b2)", fill = "Region"
  ) +
  theme(legend.position = "bottom")
```

## Dataset Summary Statistics

```{r findings}
cat("=== KEY FINDINGS ===\n\n")

cat("1. Drinking Vessel Prevalence:\n")
cat("   -", sum(cups_skulls_clean$has_drinking_vessel == TRUE, na.rm = TRUE),
    "graves (", round(sum(cups_skulls_clean$has_drinking_vessel == TRUE, na.rm = TRUE) / nrow(cups_skulls_clean) * 100, 1),
    "%) have drinking vessels recorded\n\n")

cat("2. Skeletal Remains:\n")
cat("   -", sum(cups_skulls_clean$has_remains == TRUE, na.rm = TRUE),
    "graves (", round(sum(cups_skulls_clean$has_remains == TRUE, na.rm = TRUE) / nrow(cups_skulls_clean) * 100, 1),
    "%) have human remains\n")
cat("   - Skull counts recorded in", sum(!is.na(cups_skulls_clean$skulls_number)), "graves\n\n")

cat("3. Hair Pin Prevalence:\n")
cat("   -", sum(cups_skulls_clean$has_pin == TRUE, na.rm = TRUE),
    "graves (", round(sum(cups_skulls_clean$has_pin == TRUE, na.rm = TRUE) / nrow(cups_skulls_clean) * 100, 1),
    "%) have hair pins recorded\n\n")

cat("4. Burial Type Association:\n")
burial_counts <- cups_skulls_clean %>%
  count(burial_category) %>%
  arrange(desc(n))
for (i in 1:nrow(burial_counts)) {
  cat("   -", burial_counts$burial_category[i], ":", burial_counts$n[i], "graves\n")
}
```

## Chamber Space and Burial Goods

Larger tomb chambers could accommodate more burials over time, which might predict higher skull counts, more accumulated grave goods, and thus more drinking vessels. This section examines whether chamber area correlates with skull count and drinking vessel count using Pearson correlation and simple linear regression, conducted separately for each region. Scatter plots with fitted regression lines are provided alongside the statistical results.

#### Data Availability

```{r space-data-summary}
space_analysis_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(area_of_chamber_estimated))

cat("Graves with chamber area estimated:", nrow(space_analysis_data), "\n")
cat("Graves with area AND skull count:", sum(!is.na(space_analysis_data$skulls_number)), "\n")
cat("Graves with area AND vessel count:", sum(!is.na(space_analysis_data$drinking_vessel_number)), "\n\n")

# Summary statistics for chamber area
area_stats <- summary(space_analysis_data$area_of_chamber_estimated)
kable(data.frame(Statistic = names(area_stats), Value = round(as.numeric(area_stats), 2)),
      caption = "Chamber Area Summary (m²)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

#### Skulls vs Chamber Area

```{r skulls-area-scatter, fig.width=14, fig.height=6}
skulls_area_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(area_of_chamber_estimated) & !is.na(skulls_number))

ggplot(skulls_area_data, aes(x = area_of_chamber_estimated, y = skulls_number, color = region)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_color_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  coord_cartesian(ylim = c(0, NA)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Relationship: Chamber Area vs Number of Skulls by Region",
    x = "Chamber Area Estimated (m²)",
    y = "Number of Skulls",
    color = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

```{r skulls-area-correlation}
# By region
for (reg in unique(skulls_area_data$region)) {
  region_data <- skulls_area_data %>% dplyr::filter(region == reg)
  cor_result <- cor.test(region_data$area_of_chamber_estimated, region_data$skulls_number)
  lm_result <- lm(skulls_number ~ area_of_chamber_estimated, data = region_data)

  cat("\n===", toupper(reg), "===\n")
  cat("Pearson correlation (r):", round(cor_result$estimate, 3),
      "| p-value:", round(cor_result$p.value, 4), "\n")
  print(kable(tidy(lm_result), digits = 3,
              caption = paste("Linear Regression — Skulls ~ Chamber Area:", toupper(reg))) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
}
```

#### Drinking Vessels vs Chamber Area

```{r vessels-area-scatter, fig.width=14, fig.height=6}
vessels_area_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(area_of_chamber_estimated) & !is.na(drinking_vessel_number))

ggplot(vessels_area_data, aes(x = area_of_chamber_estimated, y = drinking_vessel_number, color = region)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_color_manual(values = c(laconia = colors$laconia, messenia = colors$messenia)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Relationship: Chamber Area vs Number of Drinking Vessels by Region",
    x = "Chamber Area Estimated (m²)",
    y = "Number of Drinking Vessels",
    color = "Region"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

```{r vessels-area-correlation}
# By region
for (reg in unique(vessels_area_data$region)) {
  region_data <- vessels_area_data %>% dplyr::filter(region == reg)
  cor_result <- cor.test(region_data$area_of_chamber_estimated, region_data$drinking_vessel_number)
  lm_result <- lm(drinking_vessel_number ~ area_of_chamber_estimated, data = region_data)

  cat("\n===", toupper(reg), "===\n")
  cat("Pearson correlation (r):", round(cor_result$estimate, 3),
      "| p-value:", round(cor_result$p.value, 4), "\n")
  print(kable(tidy(lm_result), digits = 3,
              caption = paste("Linear Regression — Vessels ~ Chamber Area:", toupper(reg))) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
}
```

#### Both Skulls and Vessels vs Chamber Area

```{r both-area-scatter, fig.width=14, fig.height=6}
both_area_data <- cups_skulls_clean %>%
  dplyr::filter(!is.na(area_of_chamber_estimated) & !is.na(skulls_number) & !is.na(drinking_vessel_number))

ggplot(both_area_data, aes(x = area_of_chamber_estimated, y = skulls_number)) +
  geom_point(aes(color = drinking_vessel_number), size = 4, alpha = 0.7) +
  scale_color_gradient(name = "Drinking\nVessels", low = "#2166AC", high = "#B2182B") +
  coord_cartesian(ylim = c(0, NA)) +
  facet_wrap(~region, ncol = 1) +
  labs(
    title = "Chamber Area vs Skulls (colored by Drinking Vessels) by Region",
    x = "Chamber Area Estimated (m²)",
    y = "Number of Skulls"
  ) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## Session Info

```{r session-info}
sessionInfo()
```
